Hyperplane Elimination for Quickly Enumerating Local Optima

### 2015-09-17 Hyperplane Enumeration/main.tex
Updated to proof form

@@

Ising Spin Glasses are a type of MAX-CUT problem derived from statistical physics.
Each atom in the glass (vertex) can be assigned a spin, with the goal being to
find the set of assignments that minimize the energy between nearby atoms (edges).
Similar to Adjacent NKq, the $2D\pm J$ subset of Ising Spin Glasses can be polynomially
solved~\cite{saul:1994:spinglass}.
In this subset, the graph is defined as a square two-dimensional toroidal grid with each edge
weight chosen from $\{-1, 1\}$. Each vertex is assigned a spin from $\{-1, 1\}$ with
the energy in the glass equal to
\begin{equation}
\sum_{e_{ij} \in E} x_ie_{ij}x_j
  \label{eq-ising}
\end{equation}
where $e_{ij}$ is the weight of the edge vertex $i$ to vertex $j$. In Mk landscape terms
this type of spin glass has $M=2N$ and $k=2$.

@@

\subsection{All Optimal Orderings are Greedy}

First we show that all optimal orderings of bits are optimal orderings of packed moves.
Consider that lowering a bit in enumeration order can only decrease the number of skipped
solutions if doing so decreases the minimum dependency of a move. However, doing so may
allow the minimum dependency of some unrelated move to increase. Therefore when ordering
bits, all that matters is the order in which the last bit of each move is assigned.

% Proof
Moves can have their dependencies fixed in some permutation $O$ of all possible moves,
such that $O_0$ is the first move to have its dependencies fixed.
$D_O(i, m)$ is a function that returns how many unfixed dependencies move $m$ has after all
moves in $O$ before $i$ have been fixed. A greedy solution to this problem is one such
that $O_i$ is set to be the move that minimizes $D_O(i, m)$.

All non-greedy solutions must have some $i$ such that $D_O(i, O_i) > D_O(i, O_{i+1})$.
This is because $D_O(i, m)$ can only decrease as $i$ increases, meaning that if some $m$
has a lower value at $i$ than $O_i$, this property must hold for
some index between $i$ and when $m$ appears in $O$.
Finally, all non-greedy solutions must have some $\hat{i}$
that is the maximum $i$ value for which $D_O(i, O_i) > D_O(i, O_{i+1})$ is true.

Consider an optimal ordering $O^\star$ that is not greedy.
These criteria mean that $D_{O^\star}(\hat{i}, O_{\hat{i}}) > D_{O^\star}(\hat{i}, O_{\hat{i}+1})$.
Swapping $O_{\hat{i}}$ and $O_{\hat{i}+1}$ cannot change the minimum dependency of any other moves.
Performing the swap improves the minimum dependency of $O_{\hat{i}+1}$ more than it worsens
$O_{\hat{i}}$. Therefore, performing this swap makes $O^\star$ skip even more solutions,
contradicting the assertion that it was optimal. Therefore there cannot exist an optimal ordering that is not greedy.

@@

When Algorithm~\ref{alg-enumerate} is applied to a linear
function, each index of $move\_bin$ can contain at most 1
move, and a move is fitness improving whenever it disagrees with the global optimum.
Initially $index$ is set to $N-1$ and is decreased by at least 1 in any step after the first.
Before finding the global optimum no carry operations can occur, meaning the
cost of each iteration is equal to the amount by which $index$ is decreased.
Each step will flip a bit which disagreed with the global optimum.
Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
the global optimum.

After finding the global optimum, $index$ is set to 0. One iteration
is then spent adding a 1 to $index$ 0, which in the worst case requires \BigO{N}
carry operations. For all future iterations $index$ is a fitness-improving move
and is currently set to 1. In these iterations at least 1 carry operation must
occur, and $index$ cannot be decreased. Iteration ends when $index$ exceeds $N$
which requires at most $N$ carry operations. Therefore, Algorithm~\ref{alg-enumerate}
requires \BigO{N} time to reach termination after finding the global optimum. Combined
with initialization and the time to find the global optimum, the total complexity is \BigO{N},
which is optimal.

\subsection{$k$-Bound Separable Problems}
A $k$-bound separable problem is any problem that is composed of non-overlapping
subfunctions, each using using $k$ or fewer bits. Formally this means
$\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.
While there is no restriction on how variables are ordered in the problem,
performing reordering will ensure that
all variables which appear in the same $s_i$ are consecutively placed during enumeration.
As such it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
variables appear before $s_j$'s in enumeration ordering.

Algorithm~\ref{alg-enumerate} finds the $f_i$ with the highest $i$ which contains a
fitness-improving move and enumerates its variables until it no longer
contains an improving move. In total, enumerating $f_i$ requires $2^k$
steps. Each $f_i$ is considered sequentially until none contains an improving move,
meaning the first local optimum is found in \BigO{2^kN/k} time.

Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
by finding all local optimum in larger and larger hyperplanes. Initially $f_0$
is enumerated to find all local optima in the hyperplane with all $f_i, i>0$
fixed. This process ends when a carry operation modifies a bit of $f_1$. At
this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
$f_0$ is enumerated again. This finds all local optima in the hyperplane with $f_i, i>1$
fixed. We represent the time required to find all local optima in the hyperplane
with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
$T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
that it contains no fitness-improving moves.
Each setting in $l_i$ exposes a new hyperplane
that can contain local optima, which causes the recursive call to $T(i-1)$.
As all $|s_i| \leq k$ and $k$ is a
constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
Assuming for simplicity that all subfunctions use exactly $k$ bits,
the time required to find all local optima is
$T(N/k-1)\leq \sum_0^{N/k-1}2^kc^i<2^kc^{N/k}$.
The number of local optima in the landscape is $c^{N/k}$ meaning that Algorithm~\ref{alg-enumerate}
is within a constant factor ($2^k$) of optimal.

When looking for only $r$-bit local optima of $k$-bound separable problems,
Algorithm~\ref{alg-enumerate} continues to be within a constant of optimal.
For $r < k$ the additional moves reduce $c$ without increasing the cost
by more than a constant.
When $r \geq k$ there can only be one local optimum: the global optimum. Furthermore, as
only non-linearly related subsets of variables must be checked
for fitness improvements, no subsets can be added that are larger than $k$.
Therefore even if $r=N-1$, the number of subsets only grows at \BigO{N}.
As $c=1$ in this case, the cost of finding the single global optimum
only requires the \BigO{2^kN/k} time to find the first local optimum, which is
optimal.

Concatenated Traps is a commonly used $k$-bound separable problem
with $c=2$.  Algorithm~\ref{alg-enumerate} requires $2^k2^{N/k}$ time
to find all $2^{N/k}$ local optima in this problem. Note also that
in the degenerate case of $k=1$, a $k$-bound separable problem
is a linear function. As such, when $k=1$, $c=1$ and
\BigO{2^kN/k} = \BigO{N}.

>>>>

% Fancy proofs
\let\proof\relax
\let\endproof\relax
\usepackage{amsthm}

@@

Ising Spin Glasses are a type of MAX-CUT problem derived from statistical physics.
Each atom in the glass (vertex) can be assigned a spin, with the goal being to
find the set of assignments that minimize the energy between nearby atoms (edges).
Similar to Adjacent NKq, the $2D\pm J$ subset of Ising Spin Glasses can be polynomially
solved~\cite{saul:1994:spinglass}.
In this subset, the graph is defined as a square two-dimensional grid with
periodic boundaries such that each edge
weight is chosen from $\{-1, 1\}$. Each vertex is assigned a spin from $\{-1, 1\}$ with
the energy in the glass equal to
\begin{equation}
\sum_{e_{ij} \in E} x_ie_{ij}x_j
  \label{eq-ising}
\end{equation}
where $e_{ij}$ is the weight of the edge vertex $i$ to vertex $j$. In Mk landscape terms
this type of spin glass has $M=2N$ and $k=2$.

@@

\subsection{All Optimal Orderings are Greedy}
The quality of an enumeration ordering is determined by how many solutions
in the space can be skipped by Algorithm~\ref{alg-enumerate} using that ordering.
Each time the loop on Line~\ref{alg-enumerate-bincheck} stops with $index>0$,
$2^{index}$ solutions have been skipped. For practical purposes we will
assume that all moves are equally likely to be fitness improving.

\begin{lemma}
There exists an ordering of moves such that assigning all dependencies of
each move sequentially to the highest remaining positions in the enumeration ordering
will result in an optimal enumeration ordering.
\end{lemma}

\begin{proof}
The number of solutions skipped each time a move is fitness improving depends
on where it is placed in $move\_bin$. This position is determined by its
minimum dependency. Therefore, the placement of variables which are not
the minimum dependency of at least one move cannot affect the number of
skipped solutions. Consider an ordering of moves which is determined by
the minimum dependencies of each move in some enumeration ordering. This
ordering of moves can then be used to create an enumeration ordering which
skips the same number of solutions as the original enumeration ordering.
\end{proof}

Moves can have their dependencies assigned in some permutation $O$ of all possible moves,
such that $O_0$ is the first move to have its dependencies assigned.
$D_O(i, m)$ is a function that returns how many unassigned dependencies move $m$ has after all
moves in $O$ before $i$ have been assigned. A greedy solution to this problem is one such
that $O_i$ is set to be the move that minimizes $D_O(i, m)$.

\begin{theorem}
All optimal enumeration orderings are greedy orderings of moves.
\end{theorem}

\begin{proof}
All non-greedy solutions must have some $i$ such that $D_O(i, O_i) > D_O(i, O_{i+1})$.
This is because $D_O(i, m)$ can only decrease as $i$ increases, meaning that if some $m$
has a lower value at $i$ than $O_i$, this property must hold for
some index between $i$ and when $m$ appears in $O$.
Finally, all non-greedy solutions must have some $\hat{i}$
that is the maximum $i$ value for which $D_O(i, O_i) > D_O(i, O_{i+1})$ is true.

Consider the optimal ordering $O^\star$ that is not greedy and that has the minimum value for $\hat{i}$.
Swapping $O_{\hat{i}}^\star$ and $O_{\hat{i}+1}^\star$ cannot change the minimum dependency of any other moves.
Performing the swap causes $O_{\hat{i}}^\star$ to have the minimum dependency $O_{\hat{i}+1}^\star$ had before
the swap, while $O_{\hat{i}+1}^\star$'s minimum dependency is raised by at least
$D_{O^\star}(\hat{i}, O_{\hat{i}}^\star) - D_{O^\star}(\hat{i}, O_{\hat{i}+1}^\star)$. This value
cannot be negative meaning that we have now constructed a solution which either skips more
solutions than $O^\star$ or has a lower $\hat{i}$ than $O^\star$, contradicting our assertions.
Therefore there cannot exist an optimal ordering that is not greedy.
\end{proof}

\begin{theorem}
Not all greedy orderings are optimal enumeration orderings.
\end{theorem}

\begin{proof}
Consider a problem with move dependencies $m_0 = \{A, B\}$, $m_1 = \{C, D\}$, and $m_2 = \{C, D, E\}$.
There are two greedy orderings, $[m_0, m_1, m_2]$ and $[s_1, s_2, s_0]$. The former
skips $2^3+2^1$ solutions, while the latter skips $2^3+2^2$. This is because $m_1$ and
$m_2$ overlap. As a result, even though both a greedy, only the second is optimal.
\end{proof}

We have not been able to determine a method for finding the optimal move ordering efficiently.
Therefore we will use the greedy method as it is probabilistically optimal.

@@

\begin{theorem}
Algorithm~\ref{alg-enumerate} finds all local optima of any linear function
in \BigO{N} time.
\end{theorem}

\begin{proof}
When Algorithm~\ref{alg-enumerate} is applied to a linear
function, each index of $move\_bin$ can contain at most 1
move, and a move is fitness improving whenever it disagrees with the global optimum.
Initially $index$ is set to $N-1$ and is decreased by at least 1 in any step after the first.
Each step will flip a bit which disagreed with the global optimum.
Before finding the global optimum no carry operations can occur, meaning the
cost of each iteration is equal to the amount by which $index$ is decreased.
Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
the global optimum.

After finding the global optimum, $index$ is set to 0. One iteration
is then spent adding a 1 to $index$ 0, which in the worst case requires \BigO{N}
carry operations. For all future iterations $index$ is the position of the highest fitness-improving move
which is currently set to 1. In these iterations at least 1 carry operation must
occur, and $index$ cannot be decreased. Iteration ends when $index$ exceeds $N$
which requires at most $N$ carry operations. Therefore, Algorithm~\ref{alg-enumerate}
requires \BigO{N} time to reach termination after finding the global optimum. Combined
with initialization and the time to find the global optimum, the total complexity is \BigO{N},
which is optimal.
\end{proof}

\subsection{$k$-Bound Separable Problems}
A $k$-bound separable problem is any problem that is composed of non-overlapping
subfunctions, each using using $k$ or fewer bits. Formally this means
$\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.

\begin{theorem}
Algorithm~\ref{alg-enumerate} finds all \BigO{c^{M}} local optima of any $k$-bound separable function
in \BigO{2^kc^{M}} time.
\end{theorem}

\begin{proof}
While there is no restriction on how variables are ordered in the problem,
performing reordering will ensure that
all variables which appear in the same $s_i$ are consecutively placed during enumeration.
As such it is possible to create a numbering of $f_i$ such that if $i < j$ then all of $s_i$'s
variables appear before $s_j$'s in enumeration ordering.

Algorithm~\ref{alg-enumerate} finds the $f_i$ with the highest $i$ which contains a
fitness-improving move and enumerates its variables until it no longer
contains an improving move. In total, enumerating $f_i$ requires $2^k$
steps. Each $f_i$ is considered sequentially until none contains an improving move,
meaning the first local optimum is found in \BigO{2^kM} time.

Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
by finding all local optima in larger and larger hyperplanes. Initially $f_0$
is enumerated to find all local optima in the hyperplane with all $f_i, i>0$
fixed. This process ends when a carry operation modifies a bit of $f_1$. At
this point $f_1$ is enumerated, such that each time $f_1$ contains no improving moves
$f_0$ is enumerated again. This finds all local optima in the hyperplane with $f_i, i>1$
fixed. We represent the time required to find all local optima in the hyperplane
with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated.
$T(i) = |l_i|*T(i-1)+2^k$ where $|l_i|$ is the number of ways $f_i$ can be set such
that it contains no fitness-improving moves.
Each setting in $l_i$ exposes a new hyperplane
that can contain local optima, which causes the recursive call to $T(i-1)$.
As all $|s_i| \leq k$ and $k$ is a
constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
The time required to find all local optima is therefore
$T(M-1)\leq \sum_0^{M-1}2^kc^i<2^kc^{M}$.
The number of local optima in the landscape is \BigO{c^{M}} meaning that Algorithm~\ref{alg-enumerate}
is within a constant factor ($2^k$) of optimal.
\end{proof}

When looking for only $r$-bit local optima of $k$-bound separable problems,
Algorithm~\ref{alg-enumerate} continues to be within a constant of optimal.
For $r < k$ the additional moves reduce $c$ without increasing the cost
by more than a constant.
When $r \geq k$ there can only be one local optimum: the global optimum. Furthermore, as
only non-linearly related subsets of variables must be checked
for fitness improvements, no subsets can be added that are larger than $k$.
Therefore even if $r=N-1$, the number of subsets only grows at \BigO{N}.
As $c=1$ in this case, the cost of finding the single global optimum
only requires the \BigO{2^kM} time to find the first local optimum, which is
optimal.

Concatenated Traps is a commonly used $k$-bound separable problem
with $c=2$ and $M=N/k$.  Algorithm~\ref{alg-enumerate} requires $2^k2^{N/k}$ time
to find all $2^{N/k}$ local optima in this problem. Note also that
in the degenerate case of $k=1$, a $k$-bound separable problem
is a linear function. As such, when $k=1$, $c=1$ and
\BigO{2^kM}~=~\BigO{N}.
